### 2020-05-26 writeup/transfer.tex
sec 4

Theorem \ref{thm_main_informal} shows nearly tight bounds on the trade-off between model-shift bias and variance reduction.
%determined by the covariate shift matrix and the model shift.
The bounds get tighter and tighter as $\rho_1$ increases.
A formal version of Theorem \ref{thm_main_informal} is presented in Theorem \ref{thm_model_shift} and its proof is presented in Appendix \ref{sec_pfmain}.

Our second result applies to the set setting of many tasks with the same covariates.
%We extend the above result to any number of tasks that have the same covariates.
%In this section we consider the setting with $k$ many that have the same covariates.
Since the tasks all have the same number of datapoints and covariance matrix, the trade-off between model shift bias and variance will be captured by their task models $\set{\beta_i}_{i=1}^k$.
%For this setting, we derive solutions for the multi-task training and the transfer learning setting that match our insights qualitatively from Section \ref{sec_denoise}.
Let $B^{\star} = [\beta_1, \beta_2, \dots, \beta_k] \in\real^{p\times k}$ denote the underlying task model parameters.
We can also derive a tight trade-off between model shift bias and variance reduction for this setting.
The formal statement is stated in Theorem \ref{thm_many_tasks} and its proof can be found in Appendix \ref{app_proof_many_tasks}.

@@

We compare the test error of $\hat{\beta}_t^{\MTL}$ to that of $\hat{\beta}_t^{\STL}$.
For a simple example, we show that whether $\hat{\beta}_t^{\MTL}$ performs better than $\hat{\beta}_t^{\STL}$ is determined by the distance of the task models.
We derive a sharp threshold when positive transfer transitions to negative transfer, as a ratio between the model distance and the noise level.

@@

We illustrate the example in a synthetic setting.
We demonstrate our result with a simulation. (\todo {uses the tighter bound Proposition \ref{prop_model_shift_tight}?})
We consider a setting where $p = 200$, $n_1 = 90p$, $n_2 = 30p$.
	\todo{Fill in other params.}
We fix the target task and vary the source task, by varying the task model distance parameter $d$.
We show that Theorem \ref{thm_model_shift} predicts whether we can get positive or negative transfer.
Figure \ref{fig_model_shift_phasetrans} shows the result.

Specifically, the transition threshold is derived in the following proposition.
\begin{proposition}\label{prop_dist_transition}
	In the setting of Example \ref{ex_basic} with $\sigma_1 = \sigma_2 = \sigma$, assume that $c_1 > 110$ is a fixed constant.
	Whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between the model distance and the noise level:
	\begin{itemize}
		\item If ${d^2} < \frac {2\sigma^2} {3p} \frac{(c_1 + c_2 -1)^2}{c_1 (c_1 + c_2)(c_2 - 1)}$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If ${d^2} \ge \frac {3\sigma^2} {2p} \frac{(c_1 + c_2 -1)^2}{c_1(c_1 + c_2)(c_2 - 1)} $, then whp we have that $\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}

Here the constants $\frac23$ and $\frac32$ in this proposition (and the following ones) can be replaced with $(1+c_1^{-1/2})^{-4}$ and $(1-c_1^{-1/2})^{-4}$. In particular, these bounds becomes tighter as $c_1$ increases.

The proof of Proposition \ref{prop_dist_transition} involves two parts.
First, adding the source task has a positive effect of reducing the variance of the estimator, which scales with $n_1 = c_1 p$, the number of source task data points.
Second, the difference between task models $\beta_1$ and $\beta_2$ introduces an additional bias term, which scales with $pd^2$, the distance between $\beta_1$ and $\beta_2$.
Hence, the type of transfer is determined by the trade-off between model shift bias and the reduction of variance.
The proof can be found in Appendix \ref{app_proof_31}, which is based on our main result described later in Theorem \ref{thm_model_shift}.

\begin{figure}
	\begin{subfigure}[b]{0.31\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/model_shift_phase_transition.eps}
		\caption{Task model similarity}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.31\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/datapoints_phase_transition.eps}
		\caption{Data size}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.31\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/complementary.eps}
		\caption{Covariate shift}
	\end{subfigure}
	\caption{Comparing the test error of multi-task learning to single-task learning: we observe transitions from positive to negative transfer. (a) Section \ref{sec_dist}: Task model similarities; (b) Section \ref{sec_data}: Source task noise level; (c) Section \ref{sec_covariate}: \todo{same vs. different covariance matrices}.}
	\label{fig_model_shift_phasetrans}
\end{figure}

Consider a more general setting where the noise level $\sigma_1$ of task $1$ differs from the noise level $\sigma_2$ of task $2$.
We derive a sharp transition similar to Proposition \ref{prop_dist_transition}.

\begin{proposition}\label{prop_var_transition}
	In the setting of Example \ref{ex_basic} with $d$ being fixed but $\sigma_1$ varies, assume that $c_1 > 110$ is a fixed constant and $d^2 < \frac {2\sigma_2^2} {3p} \frac{(c_1 + c_2 -1)^2}{c_1 (c_1 + c_2)(c_2 - 1)}$, i.e. we have positive transfer when $\sigma_1=\sigma_2$. Then we derive the following transition as a parameter of $\sigma_1$:
	\begin{itemize}
		\item If $\sigma_1^2 \le p d^2 \cdot c_1 +\left[1+ \frac23\frac{(c_1 + c_2 - 1)^2}{(c_1 + c_2) (c_2 - 1)}\right]\cdot\sigma_2^2$, then whp $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\sigma_1^2 > p d^2 \cdot c_1 +\left[1+ \frac32\frac{(c_1 + c_2 - 1)^2}{(c_1 + c_2) (c_2 - 1)}\right] \cdot \sigma_2^2$, then whp $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}

\subsection{Data Size and Efficiency}\label{sec_data_size}

@@

>>>>

Theorem \ref{thm_main_informal} shows nearly tight bounds on the trade-off between model-shift bias and variance reduction.
%determined by the covariate shift matrix and the model shift.
The bounds get tighter and tighter as $\rho_1$ increases.
A formal version of Theorem \ref{thm_main_informal} is presented in Theorem \ref{thm_model_shift} and its proof is presented in Appendix \ref{sec_pfmain}.
While the general form of $\Delta_{\vari}$ and $\Delta_{\beta}$ can be quite complex, we show that they provide nice interpretation for several simple settings in the rest of the section.

Our second result applies to the setting of any number of tasks with the same covariates.
%We extend the above result to any number of tasks that have the same covariates.
%In this section we consider the setting with $k$ many that have the same covariates.
Since the tasks all have the same number of datapoints and covariance matrix, the trade-off between model shift bias and variance will be captured by their task models $\set{\beta_i}_{i=1}^k$.
%For this setting, we derive solutions for the multi-task training and the transfer learning setting that match our insights qualitatively from Section \ref{sec_denoise}.
Let $B^{\star} = [\beta_1, \beta_2, \dots, \beta_k] \in\real^{p\times k}$ denote the underlying task model parameters.
We can also derive a tight trade-off between model shift bias and variance reduction for this setting.
The formal statement is stated in Theorem \ref{thm_many_tasks} and its proof can be found in Appendix \ref{app_proof_many_tasks}.

@@

A folklore intuition since the seminal work of Caruana \cite{C97} is that task relatedness determines how well multi-task learning performs.
We formalize the intuition using our setup and technical tools.
For a simple example of two tasks, we show that whether $\hat{\beta}_t^{\MTL} \le \hat{\beta}_t^{\STL}$ is determined by $\norm{\beta_1 - \beta_2}^2$, the distance of the task models.
We derive a sharp threshold when positive transfer transitions to negative transfer, as a function of model distance $\norm{\beta_1 - \beta_2}^2$, data size $n_1$ abd $n_2$ and the noise level $\sigma$.

@@

We illustrate the example with a simulation.
We consider a setting where $p = 200$, $n_1 = 90p$, $n_2 = 30p$.
We fix the target task and vary the source task, in particular the parameter $d$ which determines $\norm{\beta_1 - \beta_2}$.
Figure \ref{fig_model_shift} shows the result.
Based on Theorem \ref{thm_model_shift}, we derive the transition threshold in the following proposition.
Define the following function
\begin{align*}
	\Phi(\rho_1, \rho_2) = \frac{(\rho_1 + \rho_2 - 1)^2}{\rho_1 (\rho_1 + \rho_2) (\rho_2 - 1)}.
\end{align*}

\begin{proposition}\label{prop_dist_transition}
	In the setting of Example \ref{ex_basic} with $\sigma_1 = \sigma_2 = \sigma$, assume that $\rho_1 > 110$ is a fixed constant.
	Whether $\te(\hat{\beta}_t^{\MTL})$ is lower than $\te(\hat{\beta}_t^{\STL})$ is determined by the ratio between the model distance and the noise level:
	\begin{itemize}
		\item If ${d^2} < \frac {2\sigma^2} {3p} \cdot \Phi(\rho_1, \rho_2)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If ${d^2} \ge \frac {3\sigma^2} {2p} \cdot \Phi(\rho_1, \rho_2)$, then whp we have that $\te(\hat{\beta}_t^{\MTL}) \ge \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}

We remark that the constants $\frac23$ and $\frac32$ in the statement above are chosen for the simplicity of presentation.
In general we can close the gap between the two constants by increasing $\rho_1$.
While the statement of Theorem \ref{thm_main_informal} only applies to the asymptotic regime of $p$ going to infinity, Proposition \ref{prop_dist_transition} explains most of the observations in Figure \ref{fig_model_shift} when $p = 200$.
%in this proposition (and the following ones) can be replaced with $(1+c_1^{-1/2})^{-4}$ and $(1-c_1^{-1/2})^{-4}$. In particular, these bounds becomes tighter as $c_1$ increases.

The proof of Proposition \ref{prop_dist_transition} involves two parts.
First, in equation \eqref{eq_te_var}, the positive variance reduction effect scales with $n_1 = \rho_1 p$, the number of source task data points.
Second, we show that the negative effect of model-shift bias scales with $pd^2$, which is the expectation of $\norm{\beta_1 - \beta_2}^2$.
The proof, which is based on Theorem \ref{thm_main_informal}, can be found in Appendix \ref{app_proof_31}. %, which is based on our main result described later in Theorem \ref{thm_model_shift}.

\begin{figure}
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/model_shift_phase_transition.eps}
		\caption{Task model similarity}
		\label{fig_model_shift}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/datapoints_phase_transition.eps}
		\caption{Data size}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.32\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/complementary.eps}
		\caption{Covariate shift}
	\end{subfigure}
	\caption{Comparing the test error of multi-task learning to single-task learning: (a) positive to negative transfer as the task model dissimilarity increases (cf. Proposition \ref{prop_dist_transition}) (b) positive to negative transfer as the source task data size increases (cf. Proposition \ref{prop_data_size}); (c) }
	\label{fig_model_shift_phasetrans}
\end{figure}

\textbf{Implications.}
We further show that multi-task learning is particular powerful when the labeled data of the source task is less noisy compared to the target task.
Consider a more general setting of Proposition \ref{prop_dist_transition} where the noise level $\sigma_1$ of task $1$ differs from the noise level $\sigma_2$ of task $2$.
We derive a sharp transition from positive to negative transfer as a parameter of $\sigma_1^2$.

\begin{proposition}\label{prop_var_transition}
	In the setting of Example \ref{ex_basic} with $d$ being fixed but $\sigma_1$ varies, assume that $\rho_1 > 110$ is a fixed constant and $d^2 < \frac {2\sigma_2^2} {3p} \cdot \Phi(\rho_1, \rho_2)$.
	We derive the following transition as a parameter of $\sigma_1^2$:
	\begin{itemize}
		\item If $\sigma_1^2 \le p d^2 \cdot c_1 +\left(1+ \frac23 \cdot \Phi(\rho_1, \rho_2)\right)\cdot\sigma_2^2$, then whp $\te(\hat{\beta}_t^{\MTL}) < \te(\hat{\beta}_t^{\STL})$.
		\item If $\sigma_1^2 > p d^2 \cdot c_1 +\left(1+ \frac32\Phi(\rho_1, \rho_2)\right) \cdot \sigma_2^2$, then whp $\te(\hat{\beta}_t^{\MTL}) > \te(\hat{\beta}_t^{\STL})$.
	\end{itemize}
\end{proposition}
As a corollary, if $\sigma_1^2 \le \sigma_2^2$, then we always get positive transfer.
The proof of Proposition \ref{prop_var_transition} is similar to Proposition \ref{prop_dist_transition}.
The details can be found in Appendix \ref{app_proof_31}.

\subsection{Data Size}\label{sec_data_size}

@@

\section{Extending the Bias-Variance Trade-off to Transfer Learning}\label{sec_taskonomy}

While our focus is on multi-task learning, we show that our results extend to transfer learning settings naturally.
We provide an analysis of the transfer function of Taskonomy (Figure 4 in \cite{ZSSGM18}) using our setup.
Specifically, the source task encoder consists of the representations learnt from one or more source tasks.
The transfer function then tries to fix the target task data to the source task encoder.

We map the procedure to our setup as follows.
First, we obtain the single-task estimator $\hat{\beta}_i$ from the source tasks, for $1\le i \le t-1$.
This forms the shared representation $B^{\star}_{t-1} = [\hat{\beta}_1,\hat{\beta}_2,\dots,\hat{\beta}_{t-1}]$.
Then, we learn the output layer $W_t$ on the target task by minimizing the following objective
\begin{align}
	g(W_t) = \bignorm{X_t B W_t - Y_t}^2.
\end{align}
After solving $W_t$, we use $\hat{\beta}_t^{\TL} = B W_t$ as the estimator for the target task.
By comparing $\te(\hat{\beta}_t^{\TL})$ to $\te(\hat{\beta}_t^{\STL})$, we observe a similar trade-off between model-shift bias and variance reduction for this setting.
%We use our tools to compare $\te(B W_t)$ to $\te(\beta_t^{\STL})$.
The formal statement is presented in Theorem \ref{prop_taskonomy} and its proof in Appendix \ref{app_proof_sec4}.
